Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-11 analysis/paper/paper.tex
AGH second revisions to first draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model},
  pdfauthor={Bjørn Peare Bartholdy; Amanda G. Henry},
  pdfkeywords={oral biofilm, starch, alpha-amylase, dental calculus},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{fancyhdr}
\fancyhf{}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyheadoffset{0pt}
\rhead{\scshape A preprint - \today}
\cfoot{\thepage}

@@

\usepackage{lineno}
\linenumbers
\usepackage{subfig}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author{Bjørn Peare Bartholdy \and Amanda G. Henry}
\date{2021-10-08}

@@

% \providecommand{\@maketitle}{}
% \renewcommand{\@maketitle}{%
%   \vbox{%
%     \hsize\textwidth
%     \linewidth\hsize
%     \vskip 0.1in
%     \@toptitlebar
%     \centering
%     {\LARGE\sc \@title\par}
%     \@bottomtitlebar
%     \textsc{A Preprint}\\
%     \vskip 0.1in
%     \def\And{%
%       \end{tabular}\hfil\linebreak[0]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \def\AND{%
%       \end{tabular}\hfil\linebreak[4]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\@author\end{tabular}%
%   \vskip 0.4in \@minus 0.1in \center{\today}   \vskip 0.2in
%   }
% }
%
% \providecommand{\maketitle}{}
% \renewcommand{\maketitle}{%
%   \par
%   \begingroup
%     \renewcommand{\thefootnote}{\fnsymbol{footnote}}
%     % for perfect author name centering
%     \renewcommand{\@makefnmark}{\hbox to \z@{$^{\@thefnmark}$\hss}}
%     % The footnote-mark was overlapping the footnote-text,
%     % added the following to fix this problem               (MK)
%     \long\def\@makefntext##1{%
%       \parindent 1em\noindent
%       \hbox to 1.8em{\hss $\m@th ^{\@thefnmark}$}##1
%     }
%     \thispagestyle{empty}
%     \@maketitle
%     \@thanks
%     %\@notice
%   \endgroup
%   \let\maketitle\relax
%   \let\thanks\relax
% }

\setstretch{2.0}
Word count: 4682

@@

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine-use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).

@@

\begin{figure}
\centering
\includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview.png}
\caption{\label{fig:protocol-fig}Overview of experiment protocol including the plate setup.}
\end{figure}

@@

All scripts and data used in the analysis are available on OSF
(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}).
More detailed protocols are available on protocols.io.

@@

\begin{figure}
\includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar} \caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/ratio-plots-1.pdf}
\caption{\label{fig:ratio-plots}Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/cor-plot-1.pdf}
\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}
\end{figure}

Pearson's \emph{r} suggests a
strong positive
correlation between the total weight of the biofilms and the total starch count
(standardised by z-score) extracted from the samples across treatments,
\emph{r} = 0.659,
90\%CI{[}0.463, 0.794{]},
p = \textless{} 0.001
(Figure \ref{fig:cor-plot}).

The same test was applied to total biofilm weight and starch count per mg
calculus (also standardised by z-score), resulting in a weak positive correlation,
\emph{r} = 0.3,
90\%CI{[}0.0618, 0.506{]},
p 0.0403
(Figure \ref{fig:cor-plot2}).

\begin{figure}
\centering
\includegraphics{../figures/cor-plot2-1.pdf}
\caption{\label{fig:cor-plot2}Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}
\end{figure}

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
diameter were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, and can reach up to 100 \(\mu\)m
in diameter, whereas wheat generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
This may also suggest that granule morphology plays a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot});
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot2});
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of
starch granules in dental calculus, and that starch granules extracted from dental
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues (2015) found individual
calculus samples to be a poor predictor of diet in a population, as many of the
consumed plants were missing from some individual samples, but were present in others.\\
Power and colleagues (2015)
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
(2015),
who found that plants producing starches 10--20 \(\mu\)m in
size were over-represented; although, the representation of granules larger than
20 \(\mu\)m is unclear.

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2019; Radini et al., 2017).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calulus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g.~cooking) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface, as well as unmineralised islands and channels exist within
dental calculus and may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.\\
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.\\
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)\\
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).

A limitation of this study
is the absence of \(\alpha\)-amylase in the model. The presence of \(\alpha\)-amylase will likely affect
the total granule counts as well as the size ratios, as smaller starches may be more
susceptible to hydrolysis
(Franco et al., 1992; Haslam, 2004); however,
the absense can also allow us to directly explore
the effect of \(\alpha\)-amylase on starch counts in future experiments,
where \(\alpha\)-amylase can be added to the model
in concentrations similar to those found in the oral cavity (Scannapieco et al., 1993).
We are able to show how absolute counts in the treatments cause a difference in
incorporation; however, this
was merely a side-effect of the difference in the number of granules in potato and
wheat solutions of the same concentration (w/v). Further research should test
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown
negligible morphological changes caused by exposure to EDTA
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that
was feasible to count in a reasonable time period, so we calculated the total
counts by extrapolating from three slide transects.
Thus, we reasonably assume that the three transects are a good representation
of the entire slide, and that the distribution of all granules on the slide is
relatively homogenous.\\
Finally, we only used native starches in the experimental procedure and the results
will likely differ for processed starches (García-Granero, 2020).
Based on the comparatively low counts obtained by Leonard
and colleagues (2015, Supplement 2), processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

The data from this study are not sufficient to confirm or deny any proposed
methods of starch incorporation; however, while we are unable to address the
mechanism(s) of starch incorporation with the data obtained in this study,
the dental calculus model presented in this study is uniquely suited to explore
these questions and may improve interpretations of dietary practices in past populations.
It can address the call for more baseline testing of biases associated with
dietary research conducted on dental calculus
(Le Moyne \& Crowther, 2021).
Our high-throughput experimental setup allows us a
higher degree of control over the factors that influence starch incorporation and
retention, such as dietary intake, differential survivability of starches,
and inter- and intra-individual variation in plaque accumulation and mineralisation.
The latter is especially difficult to control \emph{in vivo} as it is influenced by
numerous factors including genetics, diet, salivary flow, and tooth position and
morphology
(Fagernäs et al., 2021; Haffajee et al., 2009; Jepsen et al., 2011; Proctor et al., 2018; Simón-Soro et al., 2013),
as well as evolutionary differences (Yates et al., 2021).
It can also facilitate training of students and researchers on methods of
dental calculus analysis, such as starch and phytolith extraction and
identification, where it can replace the use of finite archaeological resources.

@@

This research has received funding from the European Research Council under the
European Union's Horizon 2020 research and innovation program, grant agreement
number STG--677576 (``HARVEST''). We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
We did NOT make use of Sci-Hub to access articles stuck behind a paywall,
and will, therefore, NOT acknowledge the use of Sci-Hub in this study.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model},
  pdfkeywords={oral biofilm, starch, alpha-amylase, dental calculus},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{fancyhdr}
\fancyhf{}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyheadoffset{0pt}
\rhead{\scshape A preprint - \today}
\cfoot{\thepage}

@@

\usepackage{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author{Bjørn Peare Bartholdy\(^1\), Amanda G. Henry\(^1\)}
\date{\(^1\) Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.}

@@

Word count: 4722

@@

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).

@@

\begin{figure}

{\centering \includegraphics{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}

}

\caption{Overview of experiment protocol including the plate setup.}\label{fig:protocol-fig}
\end{figure}

@@

@@

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}

}

\caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{../figures/ratio-plots-1}

}

\caption{Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}\label{fig:ratio-plots}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot-1}

}

\caption{Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}\label{fig:cor-plot}
\end{figure}

Pearson's \emph{r} suggests a
strong positive
correlation between the total weight of the biofilms and the total starch count
(standardised by z-score) extracted from the samples across treatments,
\emph{r} = 0.659,
90\%CI{[}0.463, 0.794{]},
p \textless{} 0.001
(Figure \ref{fig:cor-plot}).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot2-1}

}

\caption{Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot2}
\end{figure}

The same test was applied to total biofilm weight and starch count per mg
calculus (also standardised by z-score), resulting in a weak positive correlation,
\emph{r} = 0.3,
90\%CI{[}0.0618, 0.506{]},
p = 0.0403
(Figure \ref{fig:cor-plot2}).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
diameter were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot});
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot2}).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of
starch granules in dental calculus, and that starch granules extracted from dental
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues (2015) found individual
calculus samples to be a poor predictor of diet in a population, as many of the
consumed plants were missing from some individual samples, but were present in others.\\
Power and colleagues (2015)
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
(2015),
who found that plants producing starches 10--20 \(\mu\)m in
size were over-represented; however, the representation of granules larger than
20 \(\mu\)m in their study is unclear.

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2019; Radini et al., 2017).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calulus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface of the dental calculus, as well as unmineralised islands
and channels may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
Future study should explore how burial affects the recovery of starch from the
biofilm model.

A limitation of this study
The absence of \(\alpha\)-amylase in the model is a limitation of this study, as
the total granule counts were not subject to hydrolysis. This would likely have
reduced and affected the size ratios, as smaller starches may be more
susceptible to hydrolysis
(Franco et al., 1992; Haslam, 2004); however,
the absence can also allow us to directly explore
the effect of \(\alpha\)-amylase on starch counts in future experiments,
where \(\alpha\)-amylase can be added to the model
in concentrations similar to those found in the oral cavity (Scannapieco et al., 1993).
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and
wheat solutions of the same concentration (w/v). Further research should test
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown
negligible morphological changes caused by exposure to EDTA
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that
was feasible to count in a reasonable time period, so we calculated the total
counts by extrapolating from three slide transects.
Thus, we reasonably assume that the three transects are a good representation
of the entire slide, and that the distribution of all granules on the slide is
relatively homogeneous.\\
Finally, we only used native starches in the experimental procedure and the results
will likely differ for processed starches (García-Granero, 2020).
Based on the comparatively low counts obtained by Leonard
and colleagues (2015, Supplement 2), processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

While we are unable to sufficiently address the
mechanism(s) of starch incorporation with the data obtained in this study,
the dental calculus model presented here is uniquely suited to explore
these questions and may improve interpretations of dietary practices in past
populations. Further analyses using this model can address the call for more
baseline testing of biases associated with dietary research conducted on dental calculus
(Le Moyne \& Crowther, 2021).
Our high-throughput experimental setup allows us a
higher degree of control over the factors that influence starch incorporation and
retention, such as dietary intake, differential survivability of starches,
and inter- and intra-individual variation in plaque accumulation and mineralisation.
The latter is especially difficult to control \emph{in vivo} as it is influenced by
numerous factors including genetics, diet, salivary flow, and tooth position and
morphology
(Fagernäs et al., 2021; Haffajee et al., 2009; Jepsen et al., 2011; Proctor et al., 2018; Simón-Soro et al., 2013),
as well as evolutionary differences (Yates et al., 2021).
It can also facilitate training of students and researchers on methods of
dental calculus analysis, such as starch and phytolith extraction and
identification, where it can replace the use of finite archaeological resources.

@@

We would like to thank Dr.~Stephanie Schnorr for help with the amylase activity protocol.
We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall,
we will NOT acknowledge the use of Sci-Hub in this study.

This research has received funding from the European Research Council under the
European Union's Horizon 2020 research and innovation program, grant agreement
number STG--677576 (``HARVEST'').
